---
title: "Diorhabda parameters"
author: "Tyson Wepprich"
date: "October 30, 2017"
output: 
  html_document:
    code_folding: hide
    toc: true
    number_sections: false
    toc_depth: 2
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
theme_set(theme_bw(base_size = 14)) 
knitr::opts_chunk$set(echo = TRUE)
```

##Summary
Use growth chamber photoperiod responses for logistic regression models to account for variation in field responses to photoperiod in voltinism models.
Approximate adult oviposition curve using DDRP parameters, convert to discrete substages for voltinism model.

```{r import data}
dat <- read.csv("data/DCA_CDL_2017.csv", header = TRUE)

# convert daylength time to numeric for modeling
dat$day <- sapply(strsplit(as.character(dat$day),":"),
  function(x) {
    x <- as.numeric(x)
    x[1]+x[2]/60
    }
)
dat$group <- as.factor(dat$group)
dat$N <- dat$ndiap + dat$nrep
dat$diapause <- dat$ndiap / dat$N
dat <- dat[-which(dat$site %in% c("Parker", "St. George")), ]
dat <- droplevels.data.frame(dat)

```

## Logistic models of photoperiod response

```{r logistic regression}

agdata <- aggregate(dat$diapause,by=list(dat$site, dat$day),FUN=mean)

output=glm(diapause~day, subset=site=='Big Bend State Park',data=dat,family=binomial)

# # This makes a response variable for survival that accounts for number of eggs in each row
# # Better to use this than percent survived
# diap.var = cbind(dat$ndiap, dat$N - dat$ndiap)
# 
# # glmer is for logistic regression with random effects of mother
# # without random effects, you would use the glm function with the binomial family
# output=glm(diap.var ~ day, subset=site=='Big Bend State Park', data=dat, family=binomial)
# 
# # mod <- glmer(diap.var ~ 1 + (1|mother), data = final.data, 
# #              family = binomial)



summary(output)
anova(output,test="Chisq")
coefs =coef(output)

d1=subset(agdata,Group.1=="Big Bend State Park")

plot(d1$x~d1$Group.2, ylab="Proportion diapause",xlab="Photoperiod (hrs)", xlim=c(16,11),col="red", cex=1.5, lwd=3,pch=0)
curve(exp(coefs[1]+coefs[2]*x)/(1+exp(coefs[1]+coefs[2]*x)),col="red")

# diap.var = cbind(dat$ndiap, dat$N - dat$ndiap)

sitemods <- dat %>% 
  nest(-site) %>% 
  mutate(fit = map(data, ~glm(cbind(ndiap, N - ndiap) ~ day, data = ., family = binomial)),
         coefs = map(fit, "coefficients"),
         preds = map(fit, ~data.frame(day = seq(12, 16, length.out = 100), 
                                 pred = predict(., newdata = data.frame(day = seq(12, 16, length.out = 100)), 
                                   type = "response"))),
         plots = map2(data, preds, ~ggplot(data = .x) +
                        geom_point(aes(x = day, y = diapause)) +
                        geom_line(data = .y, aes(x = day, y = pred))))

plts <- sitemods %>% 
  mutate(plots2 = map2(site, plots, ~.y + ggtitle(.x)))
pltdat <- sitemods %>% 
  unnest(preds)
plt <- ggplot(dat, aes(x = day, y = diapause)) +
  geom_point() +
  geom_line(data = pltdat, aes(x = day, y = pred)) +
  facet_wrap(~site, ncol = 2)
plt

```


With random effects of site model.
```{r glmer}
dat$zday <- scale(dat$day)
re_mod0 <- glmer(cbind(ndiap, N - ndiap) ~ 0 + (1 + zday | site), data = dat, family = binomial)
re_mod1 <- glmer(cbind(ndiap, N - ndiap) ~ zday + (1 + zday | site), data = dat, family = binomial)
re_mod2 <- glmer(cbind(ndiap, N - ndiap) ~ zday + (1 | site), data = dat, family = binomial)

newdat <- expand.grid(list(zday = seq(-2.5, 2.5, length.out = 100),
                              site = unique(dat$site)))
newdat$pred <- predict(re_mod1, newdat, type = "response")
newdat$day <- newdat$zday * attr(dat$zday, 'scaled:scale') + attr(dat$zday, 'scaled:center')

plt <- ggplot(dat, aes(x = day, y = diapause)) +
  geom_point() +
  geom_line(data = newdat, aes(x = day, y = pred)) +
  facet_wrap(~site, ncol = 2)
plt

# fedat <- sitemods %>% 
#   unnest(preds)
  
plt2 <- plt +
  geom_line(data = pltdat, aes(x = day, y = pred), color = "red")
plt2


```

Oviposition distribution determined from DDRP estimates and skewed normal shape.

```{r}

# Galerucella skewed t distribution for oviposition
arg1 = list(mu = 97.94, sigma2 = 2241.7, shape = 3.92, nu = 9.57)
x = seq(50, 350, length.out = 1000)
y = mixsmsn:::dt.ls(x, loc = arg1$mu, sigma2 = arg1$sigma2, shape = arg1$shape, nu = arg1$nu)
newdist <- data.frame(x = x, y = y) %>% 
  arrange(x) %>% 
  mutate(CDF = cumsum(y/sum(y)))
test <- SubstageDistrib(newdist, numstage = 7, perc = .99)

# can't figure out how to fit distribution to a few specified quantiles
# just use transformed Galerucella substages
test$means <- test$means + 157



```





